---
title: "Distances to Centroids & TSP across Groups"
author: "Jake Riley"
date: `r format(Sys.Date(), "%B %d, %Y")`
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  fig.width = 12
)
```

###Set workspace
```{r}
library(tidyverse) # dplyr, ggplot
library(ggmap) # for geom_leg() call
library(dbscan) # clustering
library(geosphere) # calculates distances
library(TSP) # solves travelling salesperson problem
library(gridExtra) # combines ggplot charts
```

###Read in locations
you can change this to bring in your own data as long as the coordinates are WGS84 and in a dataframe. Use the bottom half of this chunk to clarify which columns have your coordinates, unique IDs, grouping value and the column to be used for weighting (a numeric value)
```{r echo = TRUE}
locations <-
  read.csv("medschools.csv", stringsAsFactors = FALSE)

# identify the following variables
locations$id <- locations$unit_id # the unique identifier for each site
locations$lon <- locations$lon # Y coord
locations$lat <- locations$lat # X coord
locations$value <- locations$enrollment # size of points, weight for clustering

# What does this data describe
data_title <- "Medical Schools by enrollment (2015)"

# Want to use groups?
# If yes, write "TRUE" and assign a group.
# If no, write "FALSE" and then select clustering method
use_group <- TRUE
locations$group <- locations$region # column to group by geographically

# clsutering method
cluster_method <- c(
  "ward.D", "ward.D2", "single", "complete",
  "average", "mcquitty", "median", "centroid"
)
which_cluster_method <- 1 # defaults to 1: hclust(method = "ward.D")
```

The code should run from here

###What to do about groups: ```r ifelse(use_group == T, "use original groups", "clustering algorithm")```
```{r}
if (use_group == FALSE) {
  clusters <- hclust(dist(select(locations, lat, lon)),
    method = cluster_method[which_cluster_method]
  )

  # summary(clusters$height)
  locations$group <- paste0(
    "cluster ",
    cutree(clusters, h = mean(clusters$height))
  )

  cluster_bar <-
    locations %>%
    group_by(group) %>%
    summarise(n_points = n()) %>%
    arrange(desc(n_points)) %>%
    ungroup() %>%
    mutate(ord = row_number())

  ggplot(cluster_bar) +
    geom_boxplot(aes(y = n_points, x = 1)) +
    coord_flip() +
    ggtitle(paste0(
      "Distribution of points among the ",
      nrow(cluster_bar), " clusters"
    )) +
    theme(
      axis.text.y = element_blank(),
      axis.title.y = element_blank(),
      aspect.ratio = .25
    )
} else {
  locations$group <- locations$group
}
```


Read in airports
```{r}
airports <-
  read.csv("airports.csv", stringsAsFactors = FALSE)
```


Create centers by finding the centroids of each group, then find the closest site to the centroid, then calculate distances to that central location
```{r}
regions <-
  locations %>%
  select(group) %>%
  distinct() %>%
  arrange(group)

centroid <-
  locations %>%
  select(id, group, lon, lat) %>%
  group_by(group) %>%
  mutate(
    centroid_lon = median(lon),
    centroid_lat = median(lat)
  ) %>%
  ungroup() %>%
  # get distances to centroid of region
  rowwise() %>%
  mutate(dist.mi = distGeo(
    p1 = c(lon, lat),
    p2 = c(centroid_lon, centroid_lat)
  ) / 1609.34) %>%
  ungroup() %>%
  # identify central facility
  group_by(group) %>%
  arrange(dist.mi) %>%
  mutate(ord = row_number()) %>%
  ungroup() %>%
  mutate(central = ifelse(ord == 1, "central", "Other")) %>%
  # assign central facility as navigation point (centroid_lon, centroid_lat)
  group_by(group) %>%
  arrange(ord) %>%
  mutate(
    centroid_lon = ifelse(ord == 1, lon, 0),
    centroid_lat = ifelse(ord == 1, lat, 0)
  ) %>%
  mutate(
    centroid_lon = min(centroid_lon),
    centroid_lat = max(centroid_lat)
  ) %>%
  ungroup() %>%
  # get distances for each site to central facility
  rowwise() %>%
  mutate(
    dist.mi =
      distGeo(
        p1 = c(lon, lat),
        p2 = c(centroid_lon, centroid_lat)
      ) / 1609.34 # convert from meters to miles
  ) %>%
  ungroup()
```


Generate map elements
```{r}
# map elements that repeat across maps
map_airports <- 
  geom_point(
    data = airports, aes(x = lon, y = lat),
    size = 3, alpha = .5, color = "black",
    fill = "yellow", shape = 24
  )

map_legend <- 
  theme(
    legend.position = "bottom",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white")
  )

all_states <- map_data("state")

all_counties <- map_data("county")

basemap <-
  ggplot() +
  geom_polygon(
    data = all_counties, aes(x = long, y = lat, group = group),
    color = "grey85", fill = "grey95"
  ) +
  geom_polygon(
    data = all_states, aes(x = long, y = lat, group = group),
    color = "grey60", fill = NA, size = 0.5
  )
```

Generate polygon outlines
```{r}
create_hull <- function(i) {
  group_points <- locations %>%
    filter(group == regions$group[i]) %>%
    select(lon, lat)
  
  select_hull <- 
    chull(group_points) %>% 
    c(.[1])
  
    group_points %>% 
      slice(select_hull) %>%
      mutate(group = regions$group[i])
}

group_outline <- map_dfr(1:nrow(regions), create_hull)
```


Show basemap
```{r}
# Make map
basemap +
  geom_point(
    data = locations,
    aes(lon, lat, color = factor(group), size = value), alpha = .8
  ) +
  geom_polygon(
    data = group_outline,
    aes(lon, lat,
      group = group,
      fill = factor(group)
    ), color = "grey50", alpha = .1
  ) +
  coord_map() +
  theme(
    legend.position = "none",
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white")
  ) +
  ggtitle(paste0(
    data_title, ": ",
    nrow(regions), " groups, n=", sum(locations$value), " (uses ",
    ifelse(use_group == T, "original groups", "clustering algorithm"), ")"
  ))
```

###Create maps
Create three maps within each group: 

* map showing region
* straight lines to central location
* shortest path through all points

Points are clustered within each region as a guide for the decision making process.

```{r eval = TRUE}
create_plots <- function(i) {
  region_loop <-
    centroid %>% 
    filter(group == regions$group[i]) %>%
    left_join(select(locations, id, value), by = "id")

  # clusters
  clusters_loop <- 
    hclust(
      dist(select(region_loop, lat, lon)),
      method = cluster_method[which_cluster_method]
    )

  region_loop$cluster <- 
    cutree(
      clusters_loop, 
      h = mean(clusters_loop$height)
    )

  # table(region_loop$cluster)

  # creates shortest routes
  df <- select(region_loop, lon, lat)
  tsp <- TSP(dist(df))
  tour <- solve_TSP(tsp, method = "nearest_insertion", control = list(rep = 10))
  path.tsp <- unname(tour)
  region_loop$Path <- as.integer(tour)

  # plot(lat~lon, df[path.tsp,], type="b",asp=1)

  # map elements that repeat across maps
  map_locations <- geom_point(
    data = region_loop,
    aes(
      x = lon, y = lat, size = value,
      fill = as.factor(cluster)
    ),
    alpha = .8, color = "black", shape = 21
  )

  map_location_central <-
    geom_point(
      data = filter(region_loop, central == "central"),
      aes(lon, lat),
      color = "red"
    )

  map_projection <-
    coord_map(
      projection = "mercator",
      xlim = c(min(region_loop$lon) - 1, max(region_loop$lon) + 1),
      ylim = c(min(region_loop$lat) - 1, max(region_loop$lat) + 1)
    )

  # show location of group
  baseplot <-
    ggplot() + coord_map() +
    geom_polygon(
      data = all_states, aes(x = long, y = lat, group = group),
      color = NA, fill = "grey80", size = .5
    ) +
    geom_point(data = region_loop, aes(x = lon, y = lat)) +
    # map_location_central +
    theme_nothing()

  # plots 1 & 2
  plot1 <- # map of distance to most central location
    basemap + map_projection + map_airports + # ggplot() +
    geom_leg(
      data = region_loop,
      aes(
        x = lon, xend = centroid_lon,
        y = lat, yend = centroid_lat
      ),
      color = "#E58700", size = 1, alpha = .5
    ) +
    geom_leg(
      data = region_loop,
      aes(
        x = lon, xend = centroid_lon,
        y = lat, yend = centroid_lat
      ),
      color = "black", size = .5, alpha = .5
    ) +
    map_locations + map_location_central + map_legend +
    guides(fill = FALSE) +
    labs(
      title = paste0(
        region_loop$group[1],
        ": ", nrow(region_loop), " sites (avg dist ",
        round(median(region_loop$dist.mi), 0), " mi.)"
      ), 
      size = "Size"
    )

  plot2 <- # map of shortest routes
    basemap + map_projection + map_airports +
    geom_path(
      data = region_loop[path.tsp, ],
      aes(x = lon, y = lat), color = "#E58700", size = 1, alpha = .5
    ) +
    geom_path(
      data = region_loop[path.tsp, ],
      aes(x = lon, y = lat), color = "black", size = .5, alpha = .5
    ) +
    map_locations + map_location_central + map_legend +
    guides(size = FALSE) +
    labs(
      title = paste0("(n = ", sum(region_loop$value), ")"),
      fill = "Cluster"
    )
  
  # put all plots together
  grid.arrange(baseplot, plot1, plot2, ncol = 3)
}

# for loop to generate maps
purrr::walk(1:nrow(regions), create_plots)
```

###Useful links

* http://stackoverflow.com/questions/27363653/find-shortest-path-from-x-y-coordinates-with-start-%E2%89%A0-end

* https://operatiology.wordpress.com/2014/05/31/tsp-with-latitudelongitude-coordinate-input-in-r/

for more information contact: rjake@sas.upenn.edu
